New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-07-11 paper/paper.Rmd
slicing 4.3

We continue the investigation by now slicing based on the projection $A_1$. For both models we look at a thin ($h=0.5$) slice through the center, $S_1^0$. As a first option we explore how manually changing the projection away from $A_1$ can help with understanding the boundary better. For our example we notice that $A_1$ does not contain any contribution from the second variable (`bd`), and in our illustration we will first rotate this variable into the view.

We again start by looking at the RF model, the slice $S_1^0$ shows the block structure of the model, with the third group (green, Gentoo penguins) overlapping with both of the other groups (Fig. \ref{slice1}, top-left). This is similar for the LDA model (Fig. \ref{slice1}, top-right), but again the linearity results in different boundaries and thus differences in how the classes overlap.

By rotating in the second variable we can find a view that shows three neatly separated blocks for the RF model, and we export the corresponding projection matrix $A_2$ to see the same views for the LDA model which also shows a neater separation of the classes along the boundary (see bottom row of Fig. \ref{slice1}).

Finally we look at the data projection based on $A_2$ and find that the third group (green, Gentoo penguins) is further away from the other two groups (compared to $A_1$), see Fig. \ref{proj2}.

>>>>

We continue the investigation by now slicing orthogonally to  the projection $A_1$. For both models we look at a thin ($h=0.5$) slice through the center, $S_1^0 = (0,0,0,0)$. At  first we explore how changing the projection, and thus the slice) away from $A_1$ can help with understanding the boundary better. For our example, notice that $A_1$ does not contain any contribution from the second variable (`bd`), so we will first rotate this variable into the view.

Figure \ref{slice1} shows snapshots of the exploration. The rop row is the initial projection ($A_1$ and $S_1^0$), and the bottom row is a later projection containing more of `bd` (call them $A_2$ and $S_2^0$). The columns correspond to the two models, with Rf on the left and LDA on the right.

Both models have some overlap at the center, even for this thin slice ($S_1^0$). The second slice ($S_2^0$) mostly resolves this overlap and reveals the primary differences between the models. The boundary between Adelie (blue) and Chinstrap (yellow) is very similar for both models. The boundary between Chinstrap and Gentoo (green) is where they differ.

The RF is almost straight in this view, as something we might expect from a tree model where splitting occurs on single variables only. However, it is straight in a combination of the second and third variable (`bd` and `fl`). By examining the scatterplot matrix, we can understand how this boundary was built: on each of `bd` and `fl` it is possible to cut on a value where most of the Gentoo penguins are different from the other two speciesfor any tree, and likely only one is needed. Thus the forest construction is providing a sample of trees that use one variable or the other variable to split, producing the blocky boundary on the combination of the two.

In contrast, the LDA boundary has an oblique split between the two, and roughly divides the space into three similarly sized areas. It is almost like a textbook illustration of how LDA works for two variables (2D) where by assuming the clusters have equal variance-covariance, it places a boundary half-way between the group means.

<!-- We again start by looking at the RF model, the slice $S_1^0$ shows the block structure of the model, with the third group (green, Gentoo penguins) overlapping with both of the other groups (Fig. \ref{slice1}, top-left). This is similar for the LDA model (Fig. \ref{slice1}, top-right), but again the linearity results in different boundaries and thus differences in how the classes overlap.

By rotating in the second variable we can find a view that shows three neatly separated blocks for the RF model, and we export the corresponding projection matrix $A_2$ to see the same views for the LDA model which also shows a neater separation of the classes along the boundary (see bottom row of Fig. \ref{slice1}).-->

Finally, to determine which model does the boundaries better, compare them with the $A_2$ projection of the data (Figure \ref{proj2}). Gentoo is more separated from the other two in this projection, and one can imagine that the trees in the forest has greedily grasped any one of many places to make a split to separate the group. It might be argued though the that RF boundary is cut too close to the Chinstrap species, and might lead to some unnecessary misclassification with new data. The LDA boundary is better placed for all species.

### 2022-07-11 paper/paper.tex
slicing 4.3

We continue the investigation by now slicing based on the projection
\(A_1\). For both models we look at a thin (\(h=0.5\)) slice through the
center, \(S_1^0\). As a first option we explore how manually changing
the projection away from \(A_1\) can help with understanding the
boundary better. For our example we notice that \(A_1\) does not contain
any contribution from the second variable (\texttt{bd}), and in our
illustration we will first rotate this variable into the view.

We again start by looking at the RF model, the slice \(S_1^0\) shows the
block structure of the model, with the third group (green, Gentoo
penguins) overlapping with both of the other groups (Fig. \ref{slice1},
top-left). This is similar for the LDA model (Fig. \ref{slice1},
top-right), but again the linearity results in different boundaries and
thus differences in how the classes overlap.

By rotating in the second variable we can find a view that shows three
neatly separated blocks for the RF model, and we export the
corresponding projection matrix \(A_2\) to see the same views for the
LDA model which also shows a neater separation of the classes along the
boundary (see bottom row of Fig. \ref{slice1}).

Finally we look at the data projection based on \(A_2\) and find that
the third group (green, Gentoo penguins) is further away from the other
two groups (compared to \(A_1\)), see Fig. \ref{proj2}.

>>>>

We continue the investigation by now slicing orthogonally to the
projection \(A_1\). For both models we look at a thin (\(h=0.5\)) slice
through the center, \(S_1^0 = (0,0,0,0)\). At first we explore how
changing the projection, and thus the slice) away from \(A_1\) can help
with understanding the boundary better. For our example, notice that
\(A_1\) does not contain any contribution from the second variable
(\texttt{bd}), so we will first rotate this variable into the view.

Figure \ref{slice1} shows snapshots of the exploration. The rop row is
the initial projection (\(A_1\) and \(S_1^0\)), and the bottom row is a
later projection containing more of \texttt{bd} (call them \(A_2\) and
\(S_2^0\)). The columns correspond to the two models, with Rf on the
left and LDA on the right.

Both models have some overlap at the center, even for this thin slice
(\(S_1^0\)). The second slice (\(S_2^0\)) mostly resolves this overlap
and reveals the primary differences between the models. The boundary
between Adelie (blue) and Chinstrap (yellow) is very similar for both
models. The boundary between Chinstrap and Gentoo (green) is where they
differ.

The RF is almost straight in this view, as something we might expect
from a tree model where splitting occurs on single variables only.
However, it is straight in a combination of the second and third
variable (\texttt{bd} and \texttt{fl}). By examining the scatterplot
matrix, we can understand how this boundary was built: on each of
\texttt{bd} and \texttt{fl} it is possible to cut on a value where most
of the Gentoo penguins are different from the other two speciesfor any
tree, and likely only one is needed. Thus the forest construction is
providing a sample of trees that use one variable or the other variable
to split, producing the blocky boundary on the combination of the two.

In contrast, the LDA boundary has an oblique split between the two, and
roughly divides the space into three similarly sized areas. It is almost
like a textbook illustration of how LDA works for two variables (2D)
where by assuming the clusters have equal variance-covariance, it places
a boundary half-way between the group means.

Finally, to determine which model does the boundaries better, compare
them with the \(A_2\) projection of the data (Figure \ref{proj2}).
Gentoo is more separated from the other two in this projection, and one
can imagine that the trees in the forest has greedily grasped any one of
many places to make a split to separate the group. It might be argued
though the that RF boundary is cut too close to the Chinstrap species,
and might lead to some unnecessary misclassification with new data. The
LDA boundary is better placed for all species.
